%\chapter{Linux driver}

The Linux SocketCAN driver can be build as \verb|ctucanfd.ko|.
Device configuration is specified via device tree entry with the following
structure:

\begin{verbatim}
CTU_CAN_FD_0: CTU_CAN_FD@43c30000 {
    compatible = "ctu,canfd-2";

    /* platform-dependent registers address */
    reg = <0x43c30000 0x10000>;

    /* platform-dependent interrupt configuration */
    interrupt-parent = <&intc>;
    interrupts = <0 30 4>;

    /* CAN clock source */
    clocks = <&clkc 15>;
};
\end{verbatim}

General information about SocketCAN may be found in Linux kernel documentation
\cite{doc:socketcan}. % TODO: ref
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Initialization}

The core is reset on probe (after the module is loaded or the device appears in
the device tree) and on device bring-up. Most of the initialization is done on
device bring-up (\verb|ctucan_open|), probe mainly collects device info,
initializes internal structures and checks that the device is compatible (via
DEVICE\_ID and VERSION registers).

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Mode configuration}

Following modes defined in Linux kernel are supported and may be selectively
turned on and off:
\begin{itemize}
\item Oneshot (via retransmit limit)
\item Bus error reporting
\item Listen only
\item Triple sampling
\item FD mode
\item Presume ACK
\item FD non-iso
\item \textit{loopback is not supported} (see \ref{sec:linux:unsupported-features})
\end{itemize}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Bitrate configuration}

Nominal and data bitrate setting is performed in \verb|ctucan_set_bittiming|
and \verb|ctucan_set_data_bittiming|, respectively. Although the bitrate setting
is also performed on device bring-up (\verb|ctucan_chip_start|), it is necessary
to perform it early, because the timing values calculated in the kernel might
require adjusting. This way the adjustments are observable immediately instead
of until after chip start.

The kernel timing calculation functions have only constraints on \textit{tseg1},
which is \textit{prop\_seg} + \textit{phase1\_seg} combined. \textit{tseg1} is
then split in half and stored into \textit{prog\_seg} and \textit{phase\_seg1}.
In CTU CAN FD, \textit{PROP} is 7 bits wide but \textit{PH1} only 6, so the
values must be re-distributed.
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Frame RX}

% TODO: diagram
ISR -> schedule NAPI queue ...> \verb|ctucan_rx_poll| -> loop(\verb|ctucan_rx|) -> \verb|napi_complete()|

When NAPI polling is scheduled, the \verb|ctucan_rx_poll| routine continually
processes RX frames (via \verb|ctucan_rx|) until either there is no frame
in hardware RX FIFO (in which case the NAPI queue is marked as complete) or a
maximum quota is reached.

Before the NAPI polling is scheduled, RXBNEI (and DOI) interrupts have to be
disabled (but not masked). They are re-enabled when all frames are read after
the NAPI queue is marked as complete.
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Frame TX}

% TODO: diagram

The core has several TX buffers with adjustable priority. For the purposes of
SocketCAN, they are used in FIFO fashion by rotating the priorities on each
completed TX frame. ID of next empty buffer is kept in \verb|txb_head|,
ID of the first full buffer (awaiting completion) in \verb|txb_tail| (modulo the
number of buffers).

TXB\#0 has the highest priority, each subsequent buffer has priority one lower.

\begin{figure}
    \centering
    \begin{subfigure}[t]{0.32\textwidth}
    \begin{tabular}{lc|c|c|c}
    TXB\# & 0 & 1 & 2 & 3 \\\hline
    Seq   & A & B & C &   \\
    Prio  & 7 & 6 & 5 & 4 \\\hline
          &   & T &   & H \\
    \end{tabular}
    \caption{3 frames are queued.}
    \end{subfigure}
    ~%
    \noindent
    \begin{subfigure}[t]{0.32\textwidth}
    \begin{tabular}{lc|c|c|c}
    TXB\# & 0 & 1 & 2 & 3 \\\hline
    Seq   &   & B & C &   \\
    Prio  & 4 & 7 & 6 & 5 \\\hline
          &   & T &   & H \\
    \end{tabular}
    \caption{Frame A was successfully sent and the priorities were rotated.}
    \end{subfigure}
    ~%
    \noindent
    \begin{subfigure}[t]{0.32\textwidth}
    \begin{tabular}{lc|c|c|c|c}
    TXB\# & 0 & 1 & 2 & 3 & 0' \\\hline
    Seq   & E & B & C & D &    \\
    Prio  & 4 & 7 & 6 & 5 &    \\\hline
          &   & T &   &   & H  \\
    \end{tabular}
    \cprotect\caption{2 new frames (D, E) were enqueued. Notice that the
        priorities did not have to be adjusted. \verb|txb_head| now contains the
        value 5 which indicates TXB\#0, but allows us to detect that all buffers
        are full.}
    \end{subfigure}
    \cprotect\caption{TXB priority rotation example. Empty Seq means the buffer
        is empty. Higher priority number means higher priority. H and T mark
        \verb|txb_head| and \verb|txb_tail|, respectively.}
\end{figure}

The network subsystem submits TX frames via \verb|ctucan_start_xmit|.
There should be at least one TXB in state TX\_ETY, otherwise an error is reported
and handled gracefully (the TX queue is stopped). Because the buffers are managed in FIFO fashion,
it is assumed that the buffer with ID \verb|txb_head| is empty and the frame is inserted there.
If all TX buffers are now full, the TX queue is stopped.

The driver gets to know about TX buffer state change via TXBHCI. In
this interrupt handler (\verb|ctucan_tx_interrupt|), it is again assumed that
the buffers come into a final (completed) state in FIFO order. This assumption
holds even if the transmission is aborted or retransmit limit is set.
If the next buffer is in a final state, we process it. Otherwise, we must:
\begin{enumerate}
\item Clear the TXBHCI interrupt -- because it was probably set again for the
      next buffers we were processing
\item Check the next buffer \textbf{again} -- because the TXBHCI might have been
      set just before we cleared it, but after we have checked the buffer the
      first time). If it is in a final state, we repeat the whole process.
\end{enumerate}

We also know the upper bound of the number of TX buffers to handle: the number
of frames currently queued in the controller (\verb|txb_head| $-$ \verb|txb_tail|).
No more buffers than this are tested (saves bus access for the final buffer).
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Unsupported features}
\label{sec:linux:unsupported-features}

\begin{itemize}
\item HW frame filters
\item Self-test mode (a.k.a. internal loopback)
\item Timestamp recording for RX and TX frames
\item Time-based transmission
\item Setting of error counters from userspace
\end{itemize}

There is no infrastructure for HW frame filters in Linux kernel, because SW filters
are more flexible and claimed to be sufficient performance-wise \cite{linux:socketcan}.
Thus no support for HW filtering is planned for the Linux driver.

Self-test mode, as implemented by the core (at the moment of writing), does not
meet the requirements for self-test mode in Linux. That may, however, change in
further releases.

Support for hardware-assisted frame timestamping is planned for future releases.
Kernel API for time-based transmission should be merged in v4.19 and the driver
might add support for this in later releases.
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Error handling}

\subsection{TX Frame drop}

In oneshot mode, when a TX frame is dropped (TXB ends up in TX\_ERR mode), the
\verb|tx_dropped| statistics counter is incremented. The same happens for
TX\_ABT mode, although buffer abort is never issued from the driver and hence
this should never happen.

\subsection{Bus errors}

\textsc{Error Passive}, \textsc{Error Warning} and \textsc{Bus Off} situations
are reported to userspace via appropriate generated error frames (see
\cite{linux:socketcan}).

Bus errors (situations leading to transmission of an error frame) and
Arbitration Lost events are received via BEI and ALI (if enabled via BERR mode
flag) and also reported to userspace via generated error frames.

\subsection{RX overrun}

Reporting RX FIFO overrun is likewise handled via generated error frame. However,
special care is needed when dealing with DOI. DOI must be disabled together
with RXBNE when NAPI polling is scheduled. Furthermore, when clearing the overrun flag,
both Data Overrun Flag and interrupt status must be cleared, in this order.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\endinput
- architecture based on SocketCAN
    - module
    - device probe
        - device tree
            - ...
    - netdev
        - ...

+ TXB priority rotation
+ TXHCI handling
+ RX overrun handling

+ Section: init
+ Section: settings (bitrate, mode, ...)
+ Section: RX
+ Section: TX
+ Section: errors


Netdev ops:
	.ndo_open	= ctucan_open,
	.ndo_stop	= ctucan_close,
+	.ndo_start_xmit	= ctucan_start_xmit,
	.ndo_change_mtu	= can_change_mtu,

PM ops:
    ctucan_suspend
    ctucan_resume

Platform driver:
	.probe	= ctucan_probe,
	.remove	= ctucan_remove,
	{ .compatible = "ctu,canfd-2.0", },
	{ .compatible = "ctu,ctucanfd", },

TODO:
- compatible field: version???

Documentation/ABI/testing/sysfs-class-net-statistics
